---
title: "final_project_mkdn"
author: "Zach F, Elizabeth B, & Esmeralda C"
date: "11/10/2021"
output: html_document
bibliography: final_project.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

What we still need:

- Some better summary table/visualizations for descriptive statistics

##Load Packages
```{r}
library(tidyverse)
library(rio)
library(here)
library(naniar)
library(ggridges)
library(knitr)
```

##Load Data
```{r}
demographics <- import(here("data", "Teen-Demographic-SPSS(1) copy.zip"))

phys_act <- import(here("data", "Teen-PA-SPSS(1) copy.zip"))

```

## Joined Data
Our data was split into 2 files for size/downloadable reasons so we
joined it to make one usable data file. We then wanted to characterize only select variables, hence the use of the select--> characterize functions. We could not find a way to characterize the data set as a whole without turning some variables' values to NA. W
We chose full join because we wanted all of the variables to appropriately merge the data sets.
```{r}
joined_set <- demographics %>% full_join(phys_act, by = "PID")

join1 <- joined_set %>% 
  as_tibble() %>%
  select(PID, TSEX, TSCHLTYPE, TETHRAC_RC, TGRADE, TAGE) %>% 
  characterize(c(TSEX, TSCHLTYPE, TETHRAC_RC, TGRADE, TAGE))

join2 <- joined_set %>% 
  select(PID,XTHEIGHTCM_RC, XTWEIGHTKG_RC, TETHRAC_RC, TPMPABAD, TPMPAWANT, TPMPAUPST, TPMPAIMPT, XTPREDWEEKS, XTPREDWEEKOFS, XTPREDWEEKW, TAGE)

full_joined_set <- join2 %>% full_join(join1, by = "PID")
```

## Convert "-9" to "NA" and ridding any rows with "NAs"
```{r}
final_joined <- full_joined_set %>% 
  replace_with_na_all(condition = ~.x == -9) %>% 
  na.omit(full_joined_set)
```


## Ages 13-18 (teenagers) only
```{r}
teenage_only <- final_joined %>% 
  filter(TAGE.x >= 3)
```

## Teenagers With BMI Calculated 
__teen_w_bmi_1__ is our data set to use and create visualizations BUT __we have not yet added the pivot longer and wider to this and save as object *cleaned_data* bc we want feedback on it first. Teen_w_bmi_1 is very clean as is and pivot longer helps us look at the data differently (visualizations to come!) but not necessarily 'cleaner'. We welcome feedback because we declare the absolute final data set__
```{r}
teen_w_bmi <- teenage_only %>% 
  mutate(height_meters = XTHEIGHTCM_RC / 100) %>% 
  mutate(bmi = XTWEIGHTKG_RC/(height_meters^2)) %>%
  mutate(total_pa_wk_min = XTPREDWEEKOFS + XTPREDWEEKW)

teen_w_bmi_1 <- subset(teen_w_bmi, TSEX!="Not ascertained")
teen_w_bmi_1 <- subset(teen_w_bmi_1, TSCHLTYPE!="Not ascertained")
teen_w_bmi_1 <- subset(teen_w_bmi_1, TGRADE!="Not ascertained")
```


## Grouping by School Type and Sex to Summarize Mean Age and Mean Total Weekly PA
```{r}
teen_w_bmi_1 %>%
  group_by(TSCHLTYPE, TSEX) %>%
  summarize(mean_age = mean(TAGE.x + 10), 
            mean_total_PA = mean(total_pa_wk_min)) %>% 
  kable()

# Need to make it to where we can see 5 total characters (so we can have two decimal points for the mean_total_PA)
```

## Pivot Longer with Total Exercise Per Week by When It Is Completed
```{r Pivot_Longer}

longer_teen_bmi_1 <- teen_w_bmi_1 %>% 
  pivot_longer(
    cols = starts_with("XTPREDWEEK"), 
    names_to = "MVPA_when", 
    values_to = "minutes", 
    values_drop_na = TRUE) %>% 
  select(PID, TSEX, TGRADE, TSCHLTYPE,bmi, MVPA_when, minutes) 

#played around with visualizations below. many need to be made pretty with colors and labels
  
longer_teen_bmi_1 %>%
  ggplot(aes(TGRADE, minutes))+
  geom_col(aes(fill=TSEX), 
           position= "dodge") +
  facet_wrap(~MVPA_when) +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  labs(title= "Pattern of Physical Activity by Grade and Sex",
       x= "Grade in School",
       y= "Total Number of Minutes Spend Exercising",
       fill= "Sex")
#need to order the y axis 

longer_teen_bmi_1 %>%
  ggplot(aes(TSCHLTYPE, minutes))+
  geom_col(aes(fill=TSEX), 
           position= "dodge") +
  facet_wrap(~MVPA_when) +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  labs(title= "Pattern of Physical Activity by Grade and Sex",
       x= "Type of School",
       y= "Total Number of Minutes Spend Exercising",
       fill= "Sex")


ggplot(longer_teen_bmi_1, aes(minutes, MVPA_when)) +
  geom_density_ridges(aes(fill = factor(TSEX)), 
                        alpha= 0.5)

ggplot(longer_teen_bmi_1, aes(minutes, MVPA_when)) +
  geom_density_ridges(aes(fill = factor(TGRADE)), 
                        alpha= 0.5)

ggplot(longer_teen_bmi_1, aes(minutes, TGRADE)) +
  geom_density_ridges(aes(fill = factor(MVPA_when)), 
                        alpha= 0.5)

longer_teen_bmi_1 %>%
  ggplot(aes(MVPA_when, minutes)) +
  geom_boxplot() +
  facet_wrap(~TSEX)

longer_teen_bmi_1 %>%
  ggplot(aes(MVPA_when, minutes)) +
  geom_boxplot() +
  facet_wrap(~TSCHLTYPE) +
  theme(axis.text.x = element_text(angle=90, hjust = 1)) +
  labs(title= "Pattern of Physical Activity by Type of School",
       x= "Grade in School",
       y= "Total Number of Minutes Spend Exercising")

#this one is pretty cool to show differences by sex

longer_teen_bmi_1 %>%
  ggplot(aes(MVPA_when, minutes)) +
  geom_boxplot() +
  facet_wrap(~TGRADE) 
 
#need to tidy up the labels 



longer_teen_bmi_1 %>%
  ggplot(aes(minutes, bmi)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~MVPA_when)
#so many BMI points I find it hard to interpret. but the code works!


  

longer_teen_bmi_1

#Will use longer_teen_bmi_1 for plot in future to visualize weekly exercise/location by school or gender
```

##Pivot Wider 

```{r}

# because of the types of variables and data, it doesn't make sense to use pivot wider on the actual set. so used a separate descriptive table to pivot wider instead. not adding this to the *cleaned_data* because it makes it messier.

 wider_teen_w_bmi_1 <- teen_w_bmi_1 %>% 
  group_by(TSCHLTYPE) %>%
  summarize(mean_bmi = mean(bmi)) %>%
  pivot_wider(
     names_from = TSCHLTYPE, 
     values_from = mean_bmi)
#how to rename a row?



```
# Summary Table
```{r}
summary_table <- teen_w_bmi_1 %>%
  group_by(TSCHLTYPE, TSEX) %>%
  summarize(mean_ht_m =mean(height_meters), sd_ht_m= sd(height_meters), mean_wt=mean(XTWEIGHTKG_RC), sd_wt= sd(XTWEIGHTKG_RC), mean_weekofs= mean(XTPREDWEEKOFS), sd_weekofs= sd(XTPREDWEEKOFS), mean_weekw= mean(XTPREDWEEKW), sd_weekw=sd(XTPREDWEEKW), mean_age=mean(TAGE.x +10), sd_age=sd(TAGE.x +10), mean_bmi= mean(bmi), sd_bmi=sd(bmi), mean_total_pa_wk_min = mean(total_pa_wk_min), sd_total_pa_wk_min=sd(total_pa_wk_min))





#can convert into bar chart and error bars (to show standard error and standard deviation)
#can pipe it into kable and label cols OR can change column names above

summary_table2 <- teen_w_bmi_1 %>%
  group_by(TSEX) %>%
  summarize(mean_ht_m =mean(height_meters), sd_ht_m= sd(height_meters), mean_wt=mean(XTWEIGHTKG_RC), sd_wt= sd(XTWEIGHTKG_RC), mean_weekofs= mean(XTPREDWEEKOFS), sd_weekofs= sd(XTPREDWEEKOFS), mean_weekw= mean(XTPREDWEEKW), sd_weekw=sd(XTPREDWEEKW), mean_age=mean(TAGE.x +10), sd_age=sd(TAGE.x +10), mean_bmi= mean(bmi), sd_bmi=sd(bmi), mean_total_pa_wk_min = mean(total_pa_wk_min), sd_total_pa_wk_min=sd(total_pa_wk_min)) 
    
summary_table2

#can plot it and color by school type or facet by school type 
            
```


# Visualizations

## First Visualization

```{r PA-BMI-race-ethnicity}
teen_w_bmi_1 %>% 
  ggplot(aes(bmi, total_pa_wk_min)) +
  geom_point(aes(color = TSEX), alpha = 0.5) +
  geom_smooth(method = lm, color = "black") +
  facet_wrap(~TETHRAC_RC.y) +
  theme_minimal() +
  labs(y = "Predicted Weekly MVPA (min)",
       x = "Body Mass Index",
       title = "Physical Activity and Body Mass Index",
       subtitle = "by Race/Ethnicity",
       color = "Sex",
       tag = "Figure 1")
```

## Second Visualization
```{r PA-by-school-type}
teen_w_bmi_1 %>% 
  ggplot(aes(total_pa_wk_min, TSCHLTYPE)) +
  geom_density_ridges(aes(fill = factor(TSEX)),
                      alpha = 0.7) +
  scale_fill_viridis_d(option = "magma", name = "Sex") +
  theme_ridges() +
  labs(title = "Total Physical Activity by School Type",
       y = "School Type",
       x = "Predicted Weekly MVPA (min)",
       tag = "Figure 2")
```


Down here is Zach messing around with an ANOVA
```{r}
teen_w_bmi_1 %>% 
  group_by(TSCHLTYPE) %>% 
  summarise(count = n(),
            mean = mean(total_pa_wk_min),
            sd = sd(total_pa_wk_min))

library(ggpubr)

teen_w_bmi_1 %>%
  ggboxplot(x = "TSCHLTYPE", y = "total_pa_wk_min",
            add = c("mean_se"),
            color = "TSCHLTYPE", 
            palette = c("yellow", 
                        "blue", 
                        "red", 
                        "black"),
            order = c("Another kind of school", 
                      "Home-schooled",
                      "Public school", 
                      "Private school"),
            ylab = "Minutes/Wk of PA",
            xlab = "School Type",
            title = "Weekly PA by School Type") +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

teen_w_bmi_1 %>% 
  ggline(x = "TSCHLTYPE", 
         y = "total_pa_wk_min", 
       add = c("mean_se"), 
       order = c("Another kind of school", 
                 "Home-schooled", 
                 "Public school", 
                 "Private school"),
       ylab = "Minutes/Wk of PA",
       xlab = "School Type") +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

# Differnece in MVPA by Gender

teen_w_bmi_1 %>% 
  ggline(x = "TSEX", 
         y = "total_pa_wk_min", 
       add = c("mean_se"), 
       order = c("Male",
                 "Female"),
       ylab = "Minutes/Wk of PA",
       xlab = "Gender") +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

anova_gender <- aov(total_pa_wk_min ~ TSEX, data = teen_w_bmi_1)
summary(anova_gender)

tukey_gender <- TukeyHSD(anova_gender)
summary(anova_gender)

tidy_tuke_gender <- tidy(tukey_gender, digits = 20)

tidy_tuke_gender %>% 
  kable()

# No significant difference in weekly PA between school type
anova_school <- aov(total_pa_wk_min ~ TSCHLTYPE, data = teen_w_bmi_1)
summary(anova_school)


# This shows that the younger one is, the greater the MVPA/WK regardless of grade
teen_w_bmi_1 %>% 
  ggboxplot(x = "TAGE.y", 
            y = "total_pa_wk_min",
            add = c("mean_se"),
            color = "TAGE.y", 
            palette = c("yellow", "blue", 
                        "red", "black",
                        "grey"),
            order = c("13 years old", "14 years old",
                      "15 years old", "16 years old",
                      "17 years old"),
            ylab = "Minutes/Wk of PA",
            xlab = "Age") + 
  facet_wrap(~TGRADE) +
  theme(axis.text.x = element_text(angle=90, hjust = 1))

# Significant differences in PA min/wk based upon age
anova_age <- aov(total_pa_wk_min ~ TAGE.y, data = teen_w_bmi_1)
summary(anova_age)
tukey_age <- TukeyHSD(anova_age)
# This shows full p-values. all are p < .05
print(tukey_age, digits = 20)

#Make output as a table
library(broom)
tidy_tukey <- tidy(tukey_age, digits = 20)

tidy_tukey %>% 
  kable(digits = 15)
```



\newpage