---
title: "final_project_mkdn"
author: "Zach F, Elizabeth B, & Esmeralda C"
date: "11/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Load Packages
```{r}
library(tidyverse)
library(rio)
library(here)
library(naniar)
library(ggridges)
```

##Load Data
```{r}
demographics <- import(here("data", "Teen-Demographic-SPSS(1) copy.zip"))

phys_act <- import(here("data", "Teen-PA-SPSS(1) copy.zip"))

```

## Joined Data
```{r}
joined_set <- demographics %>% full_join(phys_act, by = "PID")

join1 <- joined_set %>% 
  as_tibble() %>%
  select(PID, TSEX, TSCHLTYPE, TETHRAC_RC, TGRADE, TAGE) %>% 
  characterize(c(TSEX, TSCHLTYPE, TETHRAC_RC, TGRADE, TAGE))

join2 <- joined_set %>% 
  select(PID,XTHEIGHTCM_RC, XTWEIGHTKG_RC, TETHRAC_RC, TPMPABAD, TPMPAWANT, TPMPAUPST, TPMPAIMPT, XTPREDWEEKS, XTPREDWEEKOFS, XTPREDWEEKW, TAGE)

full_joined_set <- join2 %>% full_join(join1, by = "PID")
```

## Convert "-9" to "NA" and ridding any rows with "NAs"
```{r}
final_joined <- full_joined_set %>% 
  replace_with_na_all(condition = ~.x == -9) %>% 
  na.omit(full_joined_set)
```


## Ages 13-18 (teenagers) only
```{r}
teenage_only <- final_joined %>% 
  filter(TAGE.x >= 3)
```

## Teenagers With BMI Calculated )
__teen_w_bmi_1__ is our data set to use and create visualizations BUT __we still need to add the pivot longer and wider to this and save as object *cleaned_data* . That will be the absolute final data set__
```{r}
teen_w_bmi <- teenage_only %>% 
  mutate(height_meters = XTHEIGHTCM_RC / 100) %>% 
  mutate(bmi = XTWEIGHTKG_RC/(height_meters^2)) %>%
  mutate(total_pa_wk_min = XTPREDWEEKS + XTPREDWEEKOFS + XTPREDWEEKW)

teen_w_bmi_1 <- subset(teen_w_bmi, TSEX!="Not ascertained")
teen_w_bmi_1 <- subset(teen_w_bmi_1, TSCHLTYPE!="Not ascertained")
teen_w_bmi_1 <- subset(teen_w_bmi_1, TGRADE!="Not ascertained")
```


## Grouping by School Type and Sex to Summarize Mean Age and Mean Total Weekly PA
```{r}
teen_w_bmi_1 %>%
  group_by(TSCHLTYPE, TSEX) %>%
  summarize(mean_age = mean(TAGE.x + 10), 
            mean_total_PA = mean(total_pa_wk_min)) %>% 
  as_tibble()

# Need to make it to where we can see 5 total characters (so we can have two decimal points for the mean_total_PA)
```

## First Visualization
```{r PA-BMI-race-ethnicity}
teen_w_bmi_1 %>% 
  ggplot(aes(bmi, total_pa_wk_min)) +
  geom_point(aes(color = TSEX), alpha = 0.5) +
  geom_smooth(method = lm, color = "black") +
  facet_wrap(~TETHRAC_RC.y) +
  theme_minimal() +
  labs(y = "Predicted Weekly MVPA (min)",
       x = "Body Mass Index",
       title = "Physical Activity and Body Mass Index",
       subtitle = "by Race/Ethnicity",
       color = "Sex",
       tag = "Figure 1")
```

## Second Visualization
```{r PA-by-school-type}
teen_w_bmi_1 %>% 
  ggplot(aes(total_pa_wk_min, TSCHLTYPE)) +
  geom_density_ridges(aes(fill = factor(TSEX)),
                      alpha = 0.7) +
  scale_fill_viridis_d(option = "magma", name = "Sex") +
  theme_ridges() +
  labs(title = "Total Physical Activity by School Type",
       x = "School Type",
       y = "Predicted Weekly MVPA (min)",
       tag = "Figure 2")
```



# I added a geom_density_ridges for second viz but can change if don't like
Stacked Bar Chart showing

# also removed library janitor as we did not use it


*__!!!!EVERYTHING BELOW THIS IS OLD DRAFT OR PRACTICE CHUNKS!!!!__*
__ALL FINALIZED CODING SHOULD GO IN THE TOP PORTION - Practice can be done below this line!__


##Practice Chunk GGPLOT from ESME
Zach added line of best fit and eliminated not ascertained responses __probably ready to be moved into finalized section__
```{r}
teen_w_bmi_1 %>% 
  ggplot(aes(bmi, total_pa_wk_min)) +
  geom_point(aes(color = TSEX)) +
  geom_smooth(method = lm) +
  facet_wrap(~TETHRAC_RC.y) +
  theme_minimal() +
  labs(y = "Total Physical Activity per Week",
       x = "Body Mass Index",
       title = "Physical Activity and Body Mass Index",
       subtitle = "by Race/Ethnicity")
```

##Practice Chunk GGPLOT ZACH
this one shows counts of students age by grade
```{r}
teen_w_bmi_1 %>% 
  ggplot(aes(x = TGRADE, fill = TAGE.y)) +
  geom_bar(position = "stack") +
  scale_y_continuous(breaks = seq(0, 300, by = 25)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

##Practice Chunk GGPLOT ZACH NUMBER 2
this one shows counts of students age by grade faceted by race/ethnicity
```{r}
teen_w_bmi_1 %>% 
  ggplot(aes(x = TGRADE, fill = TAGE.y)) +
  geom_bar(position = "stack") +
  scale_y_continuous(breaks = seq(0, 300, by = 25)) +
  facet_wrap(~TETHRAC_RC.y) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Practice chunk GGPLOT ESME
```{r another-option-for-viz}
teen_w_bmi_1 %>% 
  ggplot(aes(total_pa_wk_min, TSCHLTYPE)) +
  geom_density_ridges(aes(fill = factor(TSEX)),
                      alpha = 0.7) +
  scale_fill_viridis_d(option = "magma", name = "Sex") +
  theme_ridges() +
  labs(title = "Total Physical Activity by School Type",
       x = "School Type",
       y = "Total Weekly Physical Activity (min)",
       tag = "Figure 2")

### labs() not centered...
```


## Join Data
Datasets were **joined** (demographics dataset and physical activity dataset)
```{r}
library(tidyverse)
library(rio)
library(here)
library(janitor)
demographics <- import(here("data", "Teen-Demographic-SPSS(1) copy.zip"))

phys_act <- import(here("data", "Teen-PA-SPSS(1) copy.zip"))

# joined demographics and PA-questionnaire by participant ID "PID"
joined_set <- demographics %>% full_join(phys_act, by = "PID")

#try left_join and decide which to use, based on what data/variables we want

joined_set_revise <- demographics %>% left_join(phys_act, by = "PID")
#suggestion from Prof Nese
joined_set_revise %>% 
  as_tibble() %>%
  select(TSEX, TSCHLTYPE, TETHRAC_RC) %>% 
  characterize(c(TSEX, TSCHLTYPE, TETHRAC_RC))
#select so we only see what we're characterizing 

# selecting only variables that we may use
selected_variables <- joined_set %>% 
  select(TAGE, TGRADE, TSEX, TSCHLTYPE, XTHEIGHTCM_RC, XTWEIGHTKG_RC, TETHRAC_RC, TPMPABAD, TPMPAWANT, TPMPAUPST, TPMPAIMPT, XTPREDWEEKS, XTPREDWEEKOFS, XTPREDWEEKW)

selected_variables %>% as_tibble() %>%
  characterize(c("TSEX", "TSCHLTYPE", "TETHRAC_RC"))  

#characterize race, sex, and school type 
```

## Filter and Mutate
We **filtered** to only show _high school respondents_ & then used **mutate()**
to create a column for BMI (calculated from height/weight
```{r}
highschool_only <- selected_variables %>% 
  filter(TAGE >= 14 | TAGE < 18) %>% 
  filter(!is.na(TAGE))

HS_W_BMI <- highschool_only %>% 
  mutate(Height_M = XTHEIGHTCM_RC / 100) %>% 
  mutate(BMI = XTWEIGHTKG_RC/(Height_M^2)) %>%
  mutate(total_PA = XTPREDWEEKS + XTPREDWEEKOFS + XTPREDWEEKW)
```

## Commit 3
```{r}
HS_W_BMI %>%
  group_by(TSCHLTYPE, TSEX) %>%
  summarize(mean_age = mean(TAGE), mean_total_PA = mean(total_PA, na.rm= TRUE))

longer_hs <- highschool_only %>% 
  pivot_longer(
    cols = starts_with("XTPREDWEEK"), 
    names_to = "MVPA(source)", 
    values_to = "minutes", 
    values_drop_na = TRUE)

longer_hs2 <- highschool_only %>% 
  pivot_longer(
    cols = starts_with ("TPMPA"), 
    names_to = "TP")

 wider_hs <- highschool_only %>%
    pivot_wider(
     names_from = c(TGRADE,TSEX), 
     values_from = XTPREDWEEKS,
     values_fn = list(XTPREDWEEKS = mean))
 #doesn't make sense so will change, but it does make it wider! 

#come back to this to get rid of the -9
# next up: pivot_wider and pivot_longer
```

ZACH TEST (Turns all "-9" into "NA") (This worked)
```{r}
library(naniar)
NAs_HS_W_BMI <- HS_W_BMI %>% 
  replace_with_na_all(condition = ~.x == -9)
```

ZACH TEST to get rid of all rows containing NA (this worked)
```{r}
omit_HS_W_BMI <- na.omit(NAs_HS_W_BMI)

# TRIED to use characterize here, and it did nothing
```
## Esme Visual Plot- In Progress..
```{r}
omit_HS_W_BMI %>% 
  ggplot(aes(BMI, total_PA)) +
  geom_point(aes(color = TSEX)) +
  facet_wrap(~TETHRAC_RC) +
  theme_minimal() +
  labs(y = "Total Physical Activity per Week",
       x = "Body Mass Index",
       title = "Physical Activity and Body Mass Index")

# may look for something other than BMI
```

## ZACH trying to fix characterize issues
```{r}
join1 <- joined_set %>% 
  as_tibble() %>%
  select(PID, TSEX, TSCHLTYPE, TETHRAC_RC, TGRADE) %>% 
  characterize(c(TSEX, TSCHLTYPE, TETHRAC_RC, TGRADE))

join2 <- joined_set %>% 
  select(PID,XTHEIGHTCM_RC, XTWEIGHTKG_RC, TETHRAC_RC, TPMPABAD, TPMPAWANT, TPMPAUPST, TPMPAIMPT, XTPREDWEEKS, XTPREDWEEKOFS, XTPREDWEEKW)

try_joined_set <- join2 %>% full_join(join1, by = "PID")
```
**try_joined_set** worked!!!!!

